Appendix to "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model

### 2023-04-03 src/docs_paper/appendix.Rmd
Appendix puzzle fixes and paper edits

---
title: Appendix to "Integrated nested Laplace approximations for extended latent Gaussian models with application to the Naomi HIV model"
subtitle: |
  Corresponding author: Adam Howes (`ath19@ic.ac.uk`)
output:
  pdf_document:
    toc: yes
    number_sections: yes
    keep_tex: yes
    includes:
      in_header: preamble.tex
bibliography: citations.bib
---

@@

In this section we describe more fully the simplified Naomi model considered in the main text. In doing so we draw from the supplementary material of @eaton2021naomi.

@@

Let $u$ be a generic random effect. We use use structured random effects to enable partial pooling of information across units assessed as being similar, such as neighbouring districts or adjacent age-bands. We specify the first order auto-regressive model by $u \sim \text{AR}1(\sigma, \phi)$ such that
\begin{align*}
u_1 &\sim \left( 0, \frac{1}{1 - \rho^2} \right), \\
u_i &= \rho u_{i - 1} + \epsilon_t, \quad i = 2, \ldots
\end{align*}
where $\epsilon_i \sim \mathcal{N}(0, 1)$ is Gaussian white noise, and $|\rho| < 1$ is the lag-one correlation parameter.
We use $u \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with full conditionals
$$
u_i \, | \, u_{-i} \sim \mathcal{N} \left(\frac{\sum_{j: j \sim i} u_j}{n_{\delta i}}, \frac{\sigma^2}{n_{\delta i}}\right),
$$
where $u_{-i}$ is $u$ with the $i$th unit removed, $j \sim i$ if the units are defined as adjacent, $n_{\delta i} = |\{j:j \sim i\}|$ is the total number of adjacent units, and $\sigma > 0$ is the marginal standard deviation.
We follow recommendations of @freni2018note on scaling of precision matrices, disconnected adjacency graph components, and islands.
For the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising] we write $u \sim \text{BYM}2(\sigma, \phi)$, where $u$ is comprised of a spatially structured ICAR component $v^\star$ with proportion $\phi \in (0, 1)$ and spatially unstructured IID component $w^\star$ with proportion $1 - \phi$, both scaled to have generalised variance equal to one, and $\sigma > 0$ is the marginal standard deviation such that
$$
u = \sigma \left( \sqrt\phi \cdot v^\star + \sqrt{1 - \phi} \cdot w^\star \right).
$$

@@

We assume the household survey was run according to a complex survey design where each individual $j \in U$ has non-zero probability $\pi_j \in (0, 1)$ of appearing in the sample $S \subseteq U$. Suppose we observe an outcome $y_j$ for $j \in S$.
Let $w_j = 1 / \pi_j \times 1 / \omega_j$ be design weights, where $\omega_j$ is a non-response factor, then the weighted mean
$$
\hat y = \frac{\sum_{j \in S} w_j y_j}{\sum_{j \in S} w_j}.
$$
The Kish effective sample size is
$$
m = \frac{\left(\sum_{j \in S} w_j\right)^2}{\sum_{j \in S} w_j^2}.
$$
We make these computations using the `survey` \textsc{R} package [@lumley2004analysis].

@@

```{=tex}
\begin{table}[h]
\centering
\begin{tabularx}{\textwidth}{llXX}
\toprule
& Model section & Latent field & Hyperparameter \\ 
\midrule
\ref{sec:hiv-prev} & HIV prevalence & $36 + 3n$ & 9 \\ 
\ref{sec:art-cov} & ART coverage & $36 + 3n$ & 9 \\ 
\ref{sec:hiv-inc} & HIV incidence rate & $2 + n$ & 3 \\ 
\ref{sec:anc-test} & ANC testing & $4 + 2n$ & 2 \\ 
\ref{sec:art-attend} & ART attendance & $n$ & 1 \\ 
& Total & $78 + 7n$ & 24 \\ 
\bottomrule
\end{tabularx}
\label{tab:latent-hyper-size}
\caption{The numer of latent field parameters and hyperparameters in each model section.}
\end{table}
```

In this section, we describe the hyperparameter and latent field process specification for the model.
Whereas in the main text process and likelihood specifications are written together, here we consider the likelihood equations separately in Section \ref{sec:likelihood}.
Table \ref{tab:latent-hyper-size} gives the number of latent field parameters and hyperparameters in each section of the model.
Given a country with $n$ districts, then the total number of unique district-age-sex combinations is $n \times 2 \times 17 = 34n$.
In the case of Malawi then $n = 32$ such that the total number of latent field parameters is $491$ and the total number of hyperparameters is $24$.

@@

We model HIV prevalence $\rho_{x, s, a} \in [0, 1]$ on the logit scale using the linear predictor
$$
\text{logit}(\rho_{x, s, a}) = \beta^\rho_0 + \beta_{S}^{\rho, s = \text{M}} + u^\rho_a + u_a^{\rho, s = \text{M}} + u^\rho_x + u_x^{\rho, s = \text{M}} + u_x^{\rho, a < 15} + \eta^\rho_{R_x, s, a} 
$$
where $\beta^\rho_0 \sim \mathcal{N}(0, 5)$ is an intercept term, $\beta_{s}^{\rho, s = \text{M}} \sim \mathcal{N}(0, 5)$ is the difference in logit prevalence for men compared to women, $u^\rho_a \sim \text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ are age random effects for women, $u_a^{\rho, s = \text{M}} \sim \text{AR}1(\sigma_{AS}^\rho, \phi_{AS}^\rho)$ are age random effects for the difference in logit prevalence for men compared to women age $a$, $u^\rho_x \sim \text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ are spatial random effects for women, $u_x^{\rho, s = \text{M}} \sim \text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ are spatial random effects for the difference in logit prevalence for men compared to women in district $x$, $u_x^{\rho, a < 15} \sim \text{ICAR}(\sigma_{XA}^\rho)$ are spatial random effects for the ratio of paediatric prevalence to adult women prevalence, and $\eta^\rho_{R_x, s, a}$ are fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available.
We use the prior distributions $\mathcal{N}^{+}(0, 2.5)$ for the standard deviation terms $\{\sigma_A^\rho, \sigma_{AS}^\rho, \sigma_X^\rho, \sigma_{XS}^\rho, \sigma_{XA}^\rho\}$, $\mathcal{U}(-1, 1)$ for the AR1 correlation parameters $\{\phi_A^\rho, \phi_{AS}^\rho\}$, and $\text{Beta}(0.5, 0.5)$ for the BYM2 proportion parameters $\{\phi_X^\rho, \phi_{XS}^\rho\}$.

<!-- Latent field size -->

<!-- 1 + 1 + 17 + 17 + n + n + n = 36 + 3n -->

<!-- Hyperparameter size -->

<!-- 5 + 2 + 2 -->

@@

<!-- Latent field size -->

<!-- As above -->

<!-- Hyperparameter size -->

<!-- As above -->

@@

We model HIV incidence rate $\lambda_{x, s, a} > 0$ on the log scale using the linear predictor
$$
\log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a}^\lambda, \label{eq:hiv-inc}
$$
where $\beta^\lambda_0 \sim \mathcal{N}(0, 5)$ is an intercept term proportional to the average HIV transmission rate for untreated HIV positive adults, $\beta_S^{\lambda, s = \text{M}} \sim \mathcal{N}(0, 5)$ is the log incidence rate ratio for men compared to women,
$$
\rho_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a}}, \label{eq:rho-agg}
$$
is the HIV prevalence among adults 15-49 calculated by aggregating age-specific HIV prevalences, and
$$
\alpha_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}, \label{eq:alpha-agg}
$$
is the ART coverage among adults 15-49 calculated by aggregating age-specific ART coverages.
The term $\omega$ is the average reduction in HIV transmission rate per 1% increase in population ART coverage and is fixed at $\omega = 0.7$ based on inputs to the Estimation and Projection Package (EPP) model [@eaton2019estimation], $u_x^\lambda \sim \mathcal{N}(0, \sigma^\lambda)$ with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ are IID spatial random effects, and $\eta^\lambda_{R_x, s, a}$ specify fixed log incidence rate ratios by sex and age group calculated from Spectrum model output.

@@

<!-- Latent field size -->

<!-- 1 + 1 + n + 1 + 1 = 4 + n -->

<!-- Hyperparameter size -->

<!-- 3 -->

@@

For HIV prevalence, ART coverage and recent HIV infections, denoted by $\theta \in \{\rho, \alpha, \kappa\}$, the household survey $\nu$ furnishes weighted observations $\hat \theta_{\{x\}, \{s\}, \{a\}}$ with respective Kish effective sample sizes $m^{\hat \theta}_{\{x\}, \{s\}, \{a\}}$, and observed number of cases $y^{\hat \theta}_{\{x\}, \{s\}, \{a\}} = m^{\hat \theta}_{\{x\}, \{s\}, \{a\}} \cdot \hat \theta_{\{x\}, \{s\}, \{a\}}$.
We use following three binomial working likelihoods
\begin{align*}
y^{\hat \rho}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \rho}_{\{x\}, \{s\}, \{a\}}, \rho_{\{x\}, \{s\}, \{a\}}), \quad
\rho_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a}}, \\
y^{\hat \alpha}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \alpha}_{\{x\}, \{s\}, \{a\}}, \alpha_{\{x\}, \{s\}, \{a\}}), \quad
\alpha_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}. \\
y^{\hat \kappa}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \kappa}_{\{x\}, \{s\}, \{a\}}, \kappa_{\{x\}, \{s\}, \{a\}}), \quad
\kappa_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \kappa_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}},
\end{align*}
The generalised binomial $y \sim \text{xBin}(m, p)$ is defined for $y, m \in \mathbb{R}^+$ with $y \leq m$ such that
$$
\log p(y) = \log \Gamma(m + 1) - \log \Gamma(y + 1) - \log \Gamma(m - y + 1) + y \log p + (m - y) \log(1 - p),
$$
where the gamma function $\Gamma$ is such that $\forall n \in \mathbb{N}$, $\Gamma(n) = (n - 1)!$.

@@

We include ANC testing data for the year of the most recent survey.
Let $W^\text{ANC}_{\{x\}}$ be the number of ANC clients, $X^\text{ANC}_{\{x\}}$ the number of those with ascertained status, $Y^\text{ANC}_{\{x\}}$ the number of those with positive status (either known or tested) and $Z^\text{ANC}_{\{x\}}$ the number of ANC clients already on ART prior to first ANC, such that $W^\text{ANC}_{x} \geq X^\text{ANC}_{x} \geq y^\text{ANC}_{x} \geq Z^\text{ANC}_{x}$ for all $x \in \{x\}$.
When ANC testing data are only available for part of a given year, we denote $m^\text{ANC} \in \{1, \ldots, 12\}$ the number of months of reported data reflected in counts for that year.
The observed number of HIV positive and already on ART among ANC clients is modelled by
\begin{align*}
Y_{\{x\} }^{\text{ANC}} &\sim \text{Bin} \left(X_{\{x\} }^{\text{ANC}}, \rho_{\{x\},\{15, \ldots 45\}}^{\text{ANC}}\right), \\
Z_{\{x\} }^{\text{ANC}} &\sim \text{Bin} \left(Y_{\{x\} }^{\text{ANC}}, \alpha_{\{x\},\{15, \ldots 45\}}^{\text{ANC}}\right),
\end{align*}
where predicted prevalence and ART coverage are aggregated weighted by the predicted number of pregnant women by age $\Psi_{x, a}$
\begin{align*}
\rho_{\{x\}\{a\}}^{\text{ANC}} &= \frac{\sum_{\{x\}} \sum_{\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}}}{\sum_{\{x\}} \sum_{\{a\}} \Psi_{x, a}}, \\
\alpha_{\{x\}\{a\}}^{\text{ANC}} &= \frac{\sum_{\{x\}} \sum_{\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}} \cdot \alpha_{x, a}^{\text{ANC}}}{\sum_{\{x\}} \sum_{\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}}}.
\end{align*}

@@

1. If survey data on ART coverage by age and sex are not available then we set $u_a^\alpha = 0$ and $u_{a, s = \text{M}}^\alpha = 0$ and use the average age-sex pattern of ART coverage from the Spectrum offset $\eta_{R_x, s, a}^\alpha$.
2. If no ART data (survey or ART programme) are available but data on ART coverage among ANC clients are available, the level of ART coverage is not identifiable, but spatial variation is identifiable. In this instance, overall ART coverage is determined by the Spectrum offset, and only area random effects are estimated such that $\text{logit} \left(\alpha_{x, s, a} \right) = u_x^\alpha + \eta_{R_x, s, a}^\alpha$.
3. If survey data on recent HIV infection are not included in the model, then $\beta_0^\lambda = \beta_S^{\lambda, s = \text{M}} = u_x^\lambda = 0$. The sex ratio for HIV incidence is determined by the sex incidence rate ratio from Spectrum in the same years and the incidence rate in all districts is modelled assuming the same average HIV transmission rate for untreated adults, but varies according to district estimates of HIV prevalence and ART coverage.

For the Malawi example considered in the main text, all required data are available.

@@

\texttt{u\_rho\_x} & $u^\rho_x$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
\texttt{us\_rho\_x} & & Latent & $n$ & $\mathbb{R}^{n}$ \\
\texttt{u\_rho\_xs} & $u_x^{\rho, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
\texttt{us\_rho\_xs} & & Latent & $n$ & $\mathbb{R}^{n}$ \\

@@

\texttt{u\_alpha\_x} & $u^\alpha_x$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
\texttt{us\_alpha\_x} & & Latent & $n$ & $\mathbb{R}^{n}$ \\
\texttt{u\_alpha\_xs} & $u_x^{\alpha, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
\texttt{us\_alpha\_xs} & & Latent & $n$ & $\mathbb{R}^{n}$ \\

@@

\caption{Correspondence between mathematical notation and variable names used in our \texttt{TMB} code. The total number of hyperparameters is 24, and the total number of latent field parameters is $51 + 14n$.}

@@

```{cpp, echo=FALSE, eval=FALSE, output.var="ex", code=readLines("naomi_simple.cpp")}

>>>>

---
title: Appendix to "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model"
subtitle: |
  Corresponding author: Adam Howes (`ath19@ic.ac.uk`)
output:
  pdf_document:
    toc: yes
    number_sections: yes
    keep_tex: yes
    includes:
      in_header: preamble.tex
bibliography: citations.bib
---

@@

In this section we describe the simplified Naomi model considered in the main text in greater detail.
In doing so we draw from the supplementary material of @eaton2021naomi.

@@

We use use structured random effects to enable partial pooling of information across units assessed as being similar, such as neighbouring districts or adjacent age-bands.
Let $u$ be a generic random effect, with some length.
Three types of structured random effects are used in the model.
First, we specify the first order auto-regressive model by $u \sim \text{AR}1(\sigma, \phi)$ such that
\begin{align*}
u_1 &\sim \left( 0, \frac{1}{1 - \rho^2} \right), \\
u_i &= \rho u_{i - 1} + \epsilon_t, \quad i = 2, \ldots
\end{align*}
where $\epsilon_i \sim \mathcal{N}(0, 1)$ is Gaussian white noise, $\sigma > 0$ is the marginal standard deviation, and $|\rho| < 1$ is the lag-one correlation parameter.
Second, we use $u \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with full conditionals
$$
u_i \, | \, u_{-i} \sim \mathcal{N} \left(\frac{\sum_{j: j \sim i} u_j}{n_{\delta i}}, \frac{\sigma^2}{n_{\delta i}}\right),
$$
where $u_{-i}$ is $u$ with the $i$th unit removed, $j \sim i$ if the units are defined as adjacent, $n_{\delta i} = |\{j:j \sim i\}|$ is the total number of adjacent units, and $\sigma > 0$ is the marginal standard deviation.
We follow recommendations of @freni2018note on scaling of precision matrices, disconnected adjacency graph components, and islands.
Third, for the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising], we write $u \sim \text{BYM}2(\sigma, \phi)$, where $u$ is comprised of a spatially structured ICAR component $v$ with proportion $\phi \in (0, 1)$ and spatially unstructured IID component $w$ with proportion $1 - \phi$, both scaled to have generalised variance equal to one, and $\sigma > 0$ is the marginal standard deviation such that
$$
u = \sigma \left( \sqrt\phi \cdot v + \sqrt{1 - \phi} \cdot w \right).
$$

@@

We assume the household survey was run according to a complex survey design where each individual $j \in U$ has non-zero probability $\pi_j \in (0, 1)$ of appearing in the sample $S \subseteq U$. Suppose we observe an outcome $\theta_j \in \{0, 1\}$ for $j \in S$.
Let $w_j = 1 / \pi_j \times 1 / \omega_j$ be design weights, where $\omega_j$ is a non-response factor, then the weighted mean
$$
\hat \theta = \frac{\sum_{j \in S} w_j \theta_j}{\sum_{j \in S} w_j}.
$$
We account for survey weighting in the variance via the Kish effective sample size [@kish1965survey]
$$
m^{\hat \theta} = \frac{\left(\sum_{j \in S} w_j\right)^2}{\sum_{j \in S} w_j^2}.
$$
The observed number of indicator cases is then $y^{\hat \theta} = m^{\hat \theta} \cdot \hat \theta$.
To make these computations we use the `survey` \textsc{R} package [@lumley2004analysis].

@@

```{=tex}
\begin{table}[h]
\centering
\begin{tabularx}{\textwidth}{llXX}
\toprule
& Model section & Latent field & Hyperparameter \\ 
\midrule
\ref{sec:hiv-prev} & HIV prevalence & $22 + 5n$ & 9 \\ 
\ref{sec:art-cov} & ART coverage & $25 + 5n$ & 9 \\ 
\ref{sec:hiv-inc} & HIV incidence rate & $2 + n$ & 3 \\ 
\ref{sec:anc-test} & ANC testing & $2 + 2n$ & 2 \\ 
\ref{sec:art-attend} & ART attendance & $n$ & 1 \\ 
& Total & $51 + 14n$ & 24 \\ 
\bottomrule
\end{tabularx}
\label{tab:latent-hyper-size}
\caption{The numer of latent field parameters and hyperparameters in each model section.}
\end{table}
```

In this section, we describe the hyperparameter and latent field process specification for the model.
Whereas in the main text process and likelihood specifications are written together, here we consider the likelihood equations separately in Section \ref{sec:likelihood} to follow.
Table \ref{tab:latent-hyper-size} gives the number of latent field parameters and hyperparameters in each section of the model.
In the case of Malawi then $n = 32$ such that the total number of latent field parameters is $499$ and the total number of hyperparameters is $24$.

<!-- Had previously been getting a figure of 491 somewhere -- which is 51 + 14n + 24. Something going on there? -->

@@

We model HIV prevalence $\rho_{x, s, a} \in [0, 1]$ on the logit scale using the linear predictor
$$
\text{logit}(\rho_{x, s, a}) = \beta^\rho_0 + \beta_{S}^{\rho, s = \text{M}} + u^\rho_a + u_a^{\rho, s = \text{M}} + u^\rho_x + u_x^{\rho, s = \text{M}} + u_x^{\rho, a < 15} + \eta^\rho_{R_x, s, a} 
$$
where $\beta^\rho_0 \sim \mathcal{N}(0, 5)$ is an intercept term, $\beta_{s}^{\rho, s = \text{M}} \sim \mathcal{N}(0, 5)$ is the difference in logit prevalence for men compared to women, $u^\rho_a \sim \text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ are age random effects for women, $u_a^{\rho, s = \text{M}} \sim \text{AR}1(\sigma_{AS}^\rho, \phi_{AS}^\rho)$ are age random effects for the difference in logit prevalence for men compared to women age $a$, $u^\rho_x \sim \text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ are spatial random effects for women, $u_x^{\rho, s = \text{M}} \sim \text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ are spatial random effects for the difference in logit prevalence for men compared to women in district $x$, $u_x^{\rho, a < 15} \sim \text{ICAR}(\sigma_{XA}^\rho)$ are spatial random effects for the ratio of paediatric prevalence to adult women prevalence, and $\eta^\rho_{R_x, s, a}$ are fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available.
The two BYM2 random effects are comprised of the following unit scaled patially structured $\{v_x^{\rho}, v_x^{\rho, s = \text{M}}\}$ and spatially unstructured $\{w_x^{\rho}, w_x^{\rho, s = \text{M}}\}$ components
\begin{align*}
u_x^{\rho} &= \sigma_{X}^\rho \left( \sqrt\phi_{X}^\rho \cdot v_x^{\rho} + \sqrt{1 - \phi_{X}^\rho} \cdot w_x^{\rho} \right), \\
u_x^{\rho, s = \text{M}} &= \sigma_{XS}^\rho \left( \sqrt\phi_{XS}^\rho \cdot v_x^{\rho, s = \text{M}} + \sqrt{1 - \phi_{XS}^\rho} \cdot w_x^{\rho, s = \text{M}} \right).
\end{align*}
We use half-normal priors for the standard deviation terms $\{\sigma_A^\rho, \sigma_{AS}^\rho, \sigma_X^\rho, \sigma_{XS}^\rho, \sigma_{XA}^\rho\} \sim \mathcal{N}^{+}(0, 2.5)$, uniform priors for the AR1 correlation parameters $\{\phi_A^\rho, \phi_{AS}^\rho\} \sim \mathcal{U}(-1, 1)$, and beta priors for the BYM2 proportion parameters $\{\phi_X^\rho, \phi_{XS}^\rho\} \sim \text{Beta}(0.5, 0.5)$.

<!-- Latent field size -->
<!-- 1 + 1 + 10 + 10 + n + n + n = 36 + 3n -->
<!-- Hyperparameter size -->
<!-- 5 + 2 + 2 -->

@@

<!-- Latent field size -->
<!-- 1 + 1 + 13 + 10 + n + n + n = 36 + 3n -->
<!-- Hyperparameter size -->
<!-- 5 + 2 + 2 -->

@@

We model HIV incidence rate $\lambda_{x, s, a} > 0$ on the log scale using the linear predictor
$$
\log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a}^\lambda, \label{eq:hiv-inc}
$$
where $\beta^\lambda_0 \sim \mathcal{N}(0, 5)$ is an intercept term proportional to the average HIV transmission rate for untreated HIV positive adults, $\beta_S^{\lambda, s = \text{M}} \sim \mathcal{N}(0, 5)$ is the log incidence rate ratio for men compared to women,
$$
\rho_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a}}
$$
is the HIV prevalence among adults 15-49 calculated by aggregating age-specific HIV prevalences, and
$$
\alpha_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}
$$
is the ART coverage among adults 15-49 calculated by aggregating age-specific ART coverages.
The term $\omega$ is the average reduction in HIV transmission rate per 1% increase in population ART coverage and is fixed at $\omega = 0.7$ based on inputs to the Estimation and Projection Package (EPP) model [@eaton2019estimation], $u_x^\lambda \sim \mathcal{N}(0, \sigma^\lambda)$ with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ are IID spatial random effects, and $\eta^\lambda_{R_x, s, a}$ specify fixed log incidence rate ratios by sex and age group calculated from Spectrum model output.

@@

<!-- Latent field size -->
<!-- 1 + 1 + n + 1 + 1 = 4 + n -->
<!-- Hyperparameter size -->
<!-- 3 -->

@@

For HIV prevalence, ART coverage and recent HIV infections, denoted by $\theta \in \{\rho, \alpha, \kappa\}$, the household survey $\nu$ furnishes weighted observations $\hat \theta_{\{x\}, \{s\}, \{a\}}$ with respective Kish effective sample sizes $m^{\hat \theta}_{\{x\}, \{s\}, \{a\}}$, and observed number of cases $y^{\hat \theta}_{\{x\}, \{s\}, \{a\}} = m^{\hat \theta}_{\{x\}, \{s\}, \{a\}} \cdot \hat \theta_{\{x\}, \{s\}, \{a\}}$.
We use following three binomial working likelihoods
\begin{align*}
y^{\hat \rho}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \rho}_{\{x\}, \{s\}, \{a\}}, \rho_{\{x\}, \{s\}, \{a\}}), \quad
\rho_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a}}, \\
y^{\hat \alpha}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \alpha}_{\{x\}, \{s\}, \{a\}}, \alpha_{\{x\}, \{s\}, \{a\}}), \quad
\alpha_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}, \\
y^{\hat \kappa}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \kappa}_{\{x\}, \{s\}, \{a\}}, \kappa_{\{x\}, \{s\}, \{a\}}), \quad
\kappa_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \kappa_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}},
\end{align*}
The generalised binomial $y \sim \text{xBin}(m, p)$ is defined for $y, m \in \mathbb{R}^+$ with $y \leq m$ such that
$$
\log p(y) = \log \Gamma(m + 1) - \log \Gamma(y + 1) - \log \Gamma(m - y + 1) + y \log p + (m - y) \log(1 - p),
$$
where the gamma function $\Gamma$ is such that $\forall n \in \mathbb{N}$, $\Gamma(n) = (n - 1)!$.

@@

We include ANC testing data for the year of the most recent survey.
Let $W^\text{ANC}_{\{x\}}$ be the number of ANC clients, $X^\text{ANC}_{\{x\}}$ the number of those with ascertained status, $Y^\text{ANC}_{\{x\}}$ the number of those with positive status (either known or tested) and $Z^\text{ANC}_{\{x\}}$ the number of ANC clients already on ART prior to first ANC, such that
$$
W^\text{ANC}_{x} \geq X^\text{ANC}_{x} \geq Y^\text{ANC}_{x} \geq Z^\text{ANC}_{x},
$$
for all $x \in \{x\}$.
When ANC testing data are only available for part of a given year, we denote $m^\text{ANC} \in \{1, \ldots, 12\}$ the number of months of reported data reflected in counts for that year.
The observed number of HIV positive and already on ART among ANC clients is modelled by
\begin{align*}
Y_{\{x\} }^{\text{ANC}} &\sim \text{Bin} \left(X_{\{x\} }^{\text{ANC}}, \rho_{\{x\},\{15, \ldots 45\}}^{\text{ANC}}\right), \\
Z_{\{x\} }^{\text{ANC}} &\sim \text{Bin} \left(Y_{\{x\} }^{\text{ANC}}, \alpha_{\{x\},\{15, \ldots 45\}}^{\text{ANC}}\right),
\end{align*}
where predicted prevalence and ART coverage are aggregated weighted by the predicted number of pregnant women by age $\Psi_{x, a}$
\begin{align*}
\rho_{\{x\}\{a\}}^{\text{ANC}} &= \frac{\sum_{\{x\}} \sum_{\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}}}{\sum_{\{x\}} \sum_{\{a\}} \Psi_{x, a}}, \\
\alpha_{\{x\}\{a\}}^{\text{ANC}} &= \frac{\sum_{\{x\}} \sum_{\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}} \cdot \alpha_{x, a}^{\text{ANC}}}{\sum_{\{x\}} \sum_{\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}}}.
\end{align*}

@@

1. If survey data on HIV prevalence or ART coverage by age and sex are not available then we set $u_a^\theta = 0$ and $u_{a, s = \text{M}}^\theta = 0$ and use the average age-sex pattern of from the Spectrum offset $\eta_{R_x, s, a}^\theta$. For the Malawi example considered in the main text HIV prevalence and ART coverage data is not available for those aged 65+.
2. If no ART data, either survey or ART programme, are available but data on ART coverage among ANC clients are available, the level of ART coverage is not identifiable, but spatial variation is identifiable. In this instance, overall ART coverage is determined by the Spectrum offset, and only area random effects are estimated such that $\text{logit} \left(\alpha_{x, s, a} \right) = u_x^\alpha + \eta_{R_x, s, a}^\alpha$.
3. If survey data on recent HIV infection are not included in the model, then $\beta_0^\lambda = \beta_S^{\lambda, s = \text{M}} = u_x^\lambda = 0$. The sex ratio for HIV incidence is determined by the sex incidence rate ratio from Spectrum in the same years and the incidence rate in all districts is modelled assuming the same average HIV transmission rate for untreated adults, but varies according to district estimates of HIV prevalence and ART coverage.

@@

\texttt{u\_rho\_x} & $w^\rho_x$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
\texttt{us\_rho\_x} & $v^\rho_x$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
\texttt{u\_rho\_xs} & $w_x^{\rho, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
\texttt{us\_rho\_xs} & $v_x^{\rho, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\

@@

\texttt{u\_alpha\_x} & $w^\alpha_x$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
\texttt{us\_alpha\_x} & $v^\alpha_x$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
\texttt{u\_alpha\_xs} & $w_x^{\alpha, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
\texttt{us\_alpha\_xs} & $v_x^{\alpha, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\

@@

\caption{Correspondence between mathematical notation and variable names used in our \texttt{TMB} code. The total number of hyperparameters is 24, and the total number of latent field parameters is $51 + 14n$, where $n$ is the number of districts.}

@@

```{cpp, echo=TRUE, eval=FALSE, output.var="ex", code=readLines("naomi_simple.cpp")}
```
